---
title: "Writing a client against biohttp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing a client against biohttp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(biohttp)
```

biohttp is the transport layer under a family of bio-database clients. It knows
how to make an HTTP call and report what happened. It does not know what a gene
is, and it never will.

This is what you need to know to write a client on top of it.

## The one rule

**A call returns a value. It never raises.**

That is the whole contract. You do not write `tryCatch()` around a biohttp call,
because there is nothing to catch. A DNS failure, a 503, a 200 carrying an HTML
maintenance page: all three come back as a value you branch on.

```{r}
res <- httr2::with_mocked_responses(
  list(httr2::response(
    status_code = 200,
    headers = list(`content-type` = "application/json"),
    body = charToRaw('{"symbol":"BRCA1","entrezgene":672}')
  )),
  get_json("https://mygene.info/v3", path = "gene/672", source = "MyGene")
)

res$ok
res$status
res$data$symbol
```

## What comes back

Every call returns the same eight fields.

| Field | What it holds |
| --- | --- |
| `ok` | `TRUE` only when `status` is `"ok"`. Derived, so the two cannot disagree. |
| `status` | One of the seven levels below. This is what you branch on. |
| `http` | The HTTP code, or `NA_integer_` when no response arrived. |
| `data` | The parsed body on success. |
| `source` | The label you passed, used in the message. |
| `error` | One sentence fit to show a user. Never technical. |
| `detail` | The technical cause, for a log. Never shown to a user. |
| `ts` | When the envelope was built. |

The `error` and `detail` split is deliberate. Render `error` in the interface,
write `detail` to the log, and a stack trace never reaches a user while the log
never loses one.

### The seven statuses

```{r}
STATUS_LEVELS
```

They are ordered best to worst. The three that people miss:

- **`no_data`** is an answer, not a fault. The source was reached and has
  nothing for this query. A 404 from a lookup API usually means "no such
  record", which is information.
- **`skipped`** means nothing was sent, because the host's circuit breaker was
  open. Distinguishing this from `error` is why the enum exists: a skipped call
  is not evidence about the query, only about the host.
- **`stale`** carries real data that is past its freshness window. `ok` is
  `FALSE` so it never gets cached, but the body is still there.

## Branching

Handle the cases you can act on and let the rest fall through:

```{r}
render <- function(res) {
  switch(res$status,
    ok = paste("got", length(res$data), "fields"),
    no_data = "nothing found for that query",
    skipped = "source paused, try again shortly",
    rate_limited = "slow down",
    res$error
  )
}

render(status_ok(data = list(a = 1, b = 2), source = "MyGene"))
render(status_no_data(source = "MyGene"))
render(status_error(source = "MyGene", http = 503L))
```

If you genuinely do not care why a call failed, `body_or_null()` collapses it:

```{r}
body_or_null(status_ok(data = list(n = 1)))
body_or_null(status_error(source = "MyGene"))
```

Reach for that when the failure is not actionable, not to avoid reading the
envelope.

## Building a client

A client is a thin function per endpoint. Assemble, call, parse. Notice there is
no error handling in it, because there is nothing to handle.

```{r}
mygene_query <- function(symbol, species = "human") {
  res <- get_json(
    "https://mygene.info/v3",
    path = "query",
    query = list(q = symbol, species = species),
    source = "MyGene"
  )
  if (!res$ok) {
    return(res)
  }
  # Reshape the body, and return an envelope so the caller's branching still
  # works. Never return a bare value on success and NULL on failure: that is the
  # shape this package exists to replace.
  status_ok(
    data = pluck_at(res$data, "hits", default = list()),
    source = "MyGene",
    http = res$http
  )
}
```

Two things to copy from that:

1. **Pass `source`.** It is what makes the `error` sentence say "MyGene is
   temporarily unavailable" rather than "API is temporarily unavailable".
2. **Return an envelope, not a bare value.** A client that returns `NULL` on
   failure throws away the reason, and its caller has to invent a convention.

### Credentials

Pass a token as a header, never in the URL. It is redacted from anything
printable, and it stays out of access logs:

```{r}
req <- req_defaults(
  httr2::request("https://example.org/v1"),
  headers = list(Authorization = "Bearer a-real-token")
)
# The value is not in the printed request.
any(grepl("a-real-token", capture.output(print(req)), fixed = TRUE))
```

The wrappers take `headers` too, and fold them into the cache key, so two
callers with different tokens never share a cached response.

### Rate limits

The package supplies the mechanism; you supply the numbers, because only you
know what the service asks for:

```{r, eval = FALSE}
get_json(
  "https://example.org/v1",
  path = "lookup",
  source = "Example",
  throttle = list(capacity = 10, fill_time_s = 60)
)
```

## Asking many questions at once

One question per call is the wrong unit for most clients. A gene list, a variant
list, a panel: the shape is nearly always many questions to one source, and
asking them one at a time spends the round trip over and over on a call that is
already almost entirely waiting.

`get_json_many()` takes a list of queries instead of one:

```{r, eval = FALSE}
res <- get_json_many(
  "https://mygene.info/v3",
  path = "query",
  queries = lapply(c("BRCA1", "TP53", "EGFR"), function(g) list(q = g)),
  source = "MyGene",
  throttle = list(capacity = 10, fill_time_s = 60)
)

vapply(res, function(r) r$status, character(1))
#> [1] "ok" "ok" "ok"
```

Results come back **in the order you asked**, so you can zip them onto your
inputs by position. `post_json_many()` is the same idea for a JSON or GraphQL
endpoint.

The part that pays is the cache. Entries already held are served without a
request, and only the rest go out. Ask for fifty genes when forty are warm and
ten requests leave the process. The keys are built exactly the way `get_json()`
builds its own, so a batch reuses what a single call warmed and the other way
around.

### One source, not a mix

These are for many questions to **one** source. httr2 applies `req_throttle()`
and `req_retry()` across the whole list rather than per request, which as its own
documentation puts it makes `req_perform_parallel()` "most suitable for
performing many parallel requests to the same host, rather than a mix of
different hosts". A throttled request to one host would make an unthrottled
request to another wait behind it.

Requests are grouped by host so that stays true even if you pass a mixed list,
and each host's throttle bucket stays honest. Host groups run one after another.

If what you want is a dozen *different* services answered at once, that is
process-level concurrency with a worker pool, and it belongs in your application.
This package does not manage a daemon pool and is not trying to.

### Supply a throttle

httr2's own advice is never to perform in parallel without `req_throttle()`,
because it is otherwise very easy to flood a source with simultaneous requests.
Public biological data sources are typically run on a research budget. Pass a
`throttle`.

## The circuit breaker

After a few consecutive **transport** failures, a host is skipped for a cooldown
rather than waited on. It heals itself; there is nothing to reset.

The rule that matters:

> Only a transport failure counts against a host. Any HTTP response at all,
> including a 500 and including a 200 whose body will not parse, proves the host
> is reachable and clears the count.

This is not a detail. A source that starts returning HTML error pages with a 200
is broken, but it is *up*, and taking it out of rotation would turn a partial
outage into a total one. This is the rule a hand-rolled breaker most often gets
wrong, and a large part of why the package exists.

## Caching

Successes are cached. Failures never are.

A cache that stores a failure poisons itself for the life of the process: every
later lookup serves the stored failure instead of retrying. Storing only
successes means a transient outage resolves itself the moment the source comes
back.

Memory only by default. The disk tier is opt-in, because a library should not
start writing to somebody's disk because they installed it:

| Variable | Default | What it does |
| --- | --- | --- |
| `BIOHTTP_CACHE_TTL` | 1800 | Memory tier max age, seconds |
| `BIOHTTP_CACHE_MAX_SIZE` | 256 MB | Memory tier ceiling, bytes |
| `BIOHTTP_CACHE_MAX_N` | unlimited | Memory tier ceiling, number of entries |
| `BIOHTTP_CACHE_DISK` | off | Set to `true` to add the disk tier |
| `BIOHTTP_CACHE_DIR` | `tools::R_user_dir("biohttp", "cache")` | Where the disk tier lives |
| `BIOHTTP_CACHE_DISK_TTL` | 7 days | Disk tier max age, seconds |
| `BIOHTTP_CACHE_SALT` | empty | Set per deployment |

The two ceilings bound different things, and a long-running process can blow
through one while sitting well inside the other. `BIOHTTP_CACHE_MAX_SIZE` bounds
bytes, so a server answering thousands of small responses stays under it while
holding far more entries than you meant to. Set `BIOHTTP_CACHE_MAX_N` if the
count is what you care about. Eviction is least-recently-used in both cases.

Set `BIOHTTP_CACHE_SALT` if you enable the disk tier. Without it a shared cache
directory can collide across app versions, and it leaks which queries were run
to anyone able to probe it.

If the disk directory turns out to be unwritable, the cache degrades to
memory-only rather than failing, which is what lets the same code run in a
container with no writable volume.

## GraphQL

A GraphQL error arrives inside an HTTP 200, as a top-level `errors` array, so a
2xx is not enough to call the request successful. `graphql_error()` folds both
checks into one:

```{r}
res <- status_ok(data = list(errors = list(list(message = "bad field"))))
bad <- graphql_error(res, "gnomAD")
bad$status
```

It returns `NULL` when the call genuinely succeeded, so the idiom is:

```{r, eval = FALSE}
res <- post_json(url, body = list(query = q), source = "gnomAD")
bad <- graphql_error(res, "gnomAD")
if (!is.null(bad)) {
  return(bad)
}
```

## Migrating an existing app

If your app already returns the older four-field shape, `as_legacy_envelope()`
translates for you so a migration lands as a reviewable diff:

```{r}
old <- as_legacy_envelope(status_ok(data = list(n = 1), source = "MyGene"))
str(old)
```

It is deprecated on arrival and scheduled for removal in 0.3.0. The old shape
cannot express `skipped`, and it collapses `error` and `detail` into one field.
Use it to get the migration merged, then delete the call sites.

## Testing your client

Test offline. biohttp's own suite never touches a real host, and yours should
not either.

For everything except retry, `httr2::with_mocked_responses()` is enough:

```{r, eval = FALSE}
httr2::with_mocked_responses(
  list(httr2::response(status_code = 503)),
  expect_identical(mygene_query("BRCA1")$status, "error")
)
```

Retry is the exception. Mocking replaces the transport *below* the retry loop,
so a mocked 503 arrives exactly once no matter what `max_tries` says. If you
need to test retry, run a real local server with `webfakes`, the way
`tests/testthat/test-retry.R` does.

Worth asserting in your own suite, because these are the things that quietly
regress:

- A failure is never cached.
- A 5xx does not trip the breaker.
- A token never appears in a printed request.
